Scientific Report - Summary of Research\\ Ambizione: ``Automatized compilation of sequential software to a diverse set of hardware accelerators''\\ Grant number: PZ00P2 168016\\ Period: 01.05.2018– 31.12.2020\\ Grantee: Tobias Grosser\\ \today

### 2021-04-21 paper.tex
Use an enumeratiob
deploy: be4f93ae67ef9288b8f7d0937ee421f3fcbb13ce

The focus of year three in this Ambizione project was move our polyhedral core
infrastructure into real-world compilers such as MLIR and improve its
performance. In addition, we worked on establishing backend abstractions for 
dral techniques. In the
following, we discuss four milestones we identified as important to reach these goals:
1. Fast, robust, and (provably?) correct polyhedral libraries
2. Solid polyhedral compiler infrastructure
• Modular pattern-based optimizations for CPU and GPU systems
• DeLICM: Scalar dependence removal
3. Precise cache models for fully-associative caches
4. Domain-specific optimization for climate science
The following two goals from year one remain important:
1. Solidify and formalize our underlying models
2. Effective accelerator offloading for large-scale programs

@@

\subsection{Publications (published)}
\begin{itemize}
	\item \bibentry{chelini2021mlir}
	\item \bibentry{10.1145/3437801.3441613}
	\item \bibentry{grosser2020fast}
	\item \bibentry{chelini2020automatic}
	\item \bibentry{khan2020polyhedral}
	\item \bibentry{schueki2020llhd}
	\item \bibentry{kurth2020mixed}
	\item \bibentry{kourtis2020compiling}
\end{itemize}
\subsection{Publications (under review)}
\begin{itemize}
	\item \bibentry{gysi2020ooc}
	\item \bibentry{fehr2021}
	\item \bibentry{anurudth2021}
	\item \bibentry{theo2021}
	\item \bibentry{pichanatan2021}
	\item \bibentry{copic2021}
\end{itemize}

>>>>

The focus of year three in this Ambizione project was move our polyhedral core
infrastructure into real-world compilers (e.g, LLVM/MLIR) and improve its
performance and scalability. In addition, we worked on establishing backends
for targetting accelerators such as GPUs and FPGAs. Finally, we worked on
performance modelling. As mentioned in the year 2 report, our focus moved from
Julia to MLIR as a compiler and programming language infrastructure.

In the following, we discuss four milestones we identified as important to
reach these goals:

\begin{enumerate}
\item Fast, robust, and correct polyhedral libraries
\item Solid polyhedral compiler infrastructure
\begin{itemize}
\item  Modular pattern-based optimizations for CPU and GPU systems
\item 
DeLICM: Scalar dependence removal
\end{itemize}
\item Domain-specific optimization for climate science
\end{enumerate}

\subsection{Fast, robust, and correct polyhedral libraries}
Presburger arithmetic provides the mathematical core for the polyhedral
compilation techniques that drive analytical cache models, loop optimization
for ML and HPC, formal verification, and even hardware design.  Polyhedral
compilation is widely regarded as being slow due to the potentially high
computational cost of the underlying Presburger libraries. Researchers
typically use these libraries as powerful black-box tools, but a lack of
internal documentation and decade-old C implementations hold back broader
performance-optimization efforts. With FPL, we introduced a new library for
Presburger arithmetic for the MLIR compiler built from the ground up in modern
C++. We carefully documented its internal algorithmic foundations, use
lightweight C++ data structures to minimize memory management costs, and deployed
transprecision computing across the entire library to effectively exploit
machine integers and vector instructions. On a newly-developed comprehensive
benchmark suite for Presburger arithmetic, we show a 5.25x speedup in total
runtime over the state-of-the-art library isl in its default configuration and
3.14x when over a variant of isl optimized with element-wise transprecision
computing. We expect that the availability of a well-documented and fast
Presburger library will accelerate the adoption of polyhedral compilation
techniques in production compilers.

The central compnents of this library have already been upstreamed into the
open-source LLVM/MLIR project and are usable by the wider research and 
industrial community.

The work in this subproject has been published in \citet{grosser2020fast} and
\citet{pichanatan2021}.

\subsection{Domain-specific optimization for climate science}

Most compilers have a single core intermediate representation (IR) (e.g., LLVM)
sometimes complemented with vaguely-defined IR-like data structures. This IR is
commonly low-level and close to machine instructions. As a result,
optimizations relying on domain-specific information are either not possible or
require complex analysis to recover the missing information. In contrast,
multi-level rewriting instantiates a hierarchy of dialects (IRs), lowers
programs level-by-level, and performs code transformations at the most suitable
level. We demonstrate the effectiveness of this approach for the weather and
climate domain. In particular, we develop a prototype compiler and design
stencil- and GPU-specific dialects based on a set of newly introduced design
principles. We find that two domain-specific optimizations (500 lines of code)
realized on top of LLVM’s extensible MLIR compiler infrastructure suffice to
outperform state-of-the-art solutions. In essence, multi-level rewriting
promises to herald the age of specialized compilers composed from domain- and
target-specific dialects implemented on top of a shared infrastructure.

The GPU backend of the open-earth compiler has already been upstreamed into
the open-source LLVM/MLIR project and is useable by the wider research
and industrial community.

The work in this subproject has been published in \citet{gysi2020ooc}

@@

\subsection{Publications (published)}
\begin{enumerate}
	\item \bibentry{chelini2021mlir}
	\item \bibentry{10.1145/3437801.3441613}
	\item \bibentry{grosser2020fast}
	\item \bibentry{chelini2020automatic}
	\item \bibentry{khan2020polyhedral}
	\item \bibentry{schueki2020llhd}
	\item \bibentry{kurth2020mixed}
	\item \bibentry{kourtis2020compiling}
\end{enumerate}
\subsection{Publications (under review)}
\begin{enumerate}
	\item \bibentry{gysi2020ooc}
	\item \bibentry{fehr2021}
	\item \bibentry{anurudth2021}
	\item \bibentry{theo2021}
	\item \bibentry{pichanatan2021}
	\item \bibentry{copic2021}
\end{enumerate}
